A Survey of Quantitative Descriptions of Molecular Structure

### 2012-07-21 paper.tex
Addressed some of the review comments

Methods that aim to predict chemical and biological properties
generally require a numerical description of chemical structures. Such
numerical forms range from a set of 3D coordinates (which coupled with
appropriate atom types, is sufficient for methods such as quantum
mechanical (QM) approaches and docking to more abstract numerical
descriptions derived from 2D or 3D representations which can be useful
in statistical approaches. It is now possible to evaluate thousands of
numerical descriptors of chemical structure. As will be discussed
later, many of these descriptors are closely related or capture the
same information, allowing one to be substituted for another. The
choice of descriptors is a well-known problem and given a large
collection of them, approaches to identify a suitable subset have been
discussed extensively in the
literature~\cite{Miller:2002aa,Kohavi:1997gf}.

In addition to there being many possible descriptors defined in the
literature, there are also multiple implementations of a give
descriptor. These implementations are available in the form of
libraries (which require one to write a program to use them) or
complete applications (graphical user interface (GUI) or command
line). As a result, not only must one choose one or more descriptors
that are relevant to the problem at hand, but one must be concerned
about how they are calculated and whether such a calculation can be
reproduced across different implementations of those descriptors. It
is easy to understand \emph{why} two implementations of the same
descriptor can lead to different results - the primary reasons being
that the chemistry model that underlies the framework or toolkit used
to implement the descriptor is different. For example, a descriptor
that calculates the number of aromatic atoms may be implemented using
two toolkits with differing aromaticity models, and hence it is
possible that the values generated by the two implementations will
differ. Other sources of differences include parameters that may be
involved in the descriptor calculation and reference data values (such
as atomic radii, electronegativity values) that are employed during
descriptor calculation. While most implementations will employ the
same data sources for standard concepts (e.g., atomic weights), slight
differences in these types of input data can lead to differences in
the final descriptor value~\cite{Guha:2006ac}. As a result, in most
cases, two implementations of a descriptor do not usually give the
exact same value, though they are usually quite similar. Explicitly
explaining the differences may or may not be possible (it is usually
more difficult in the cases of commercial implementations for which
source code is not available)~\cite{Spjuth2010}.

@@

@@

Another important class of geometrical descriptors are those that
describe molecular shape. This property is a key factor in ligand-receptor interactions and is
an important approach to virtual screening. A variety of descriptors
have been devised to characterize molecular shape. The use of
overlapping atom centered Gaussians to represent molecular shape
\cite{Grant:1995aa} represents a significant computational advantage
over the traditional analytical approach of intersecting hard spheres,
supporting extremely rapid shape comparisons \cite{Grant:1999vn}. An
alternative approach described by Ritchie et al \cite{Ritchie:1999kx}
employed a spherical harmonic based representation of surfaces, that
in contrast to the Gaussian approach allows one to generate
characterizations of the surface and hence the shape at varying
resolutions. Another approach to multi-resolution shape
representations is the use of $\alpha$ shapes
\cite{Edelsbrunner:1994aa}, which represent a parameterization of the
convex hull derived from the molecular coordinates. This approach has
been employed to characterize shape and volume \cite{Wilson:2009ys}
though it is more computationally intensive than the methods mentioned
previously. In addition to functional representations of molecular
shape, a number of simpler formulations have been defined. For
example, the ``shape signature'' approach described by Zauhar et al
\cite{Zauhar:2003fk} is a novel application of a ray tracing method
that generates a description of the molecular shape. Ballester et al
\cite{Ballester:2007aa} described the Ultrafast Shape Recognition
(USR) approach that characterized molecular shape in terms of the
distribution of distances between the atoms and a set of pre-defined
points. This procedure characterizes the shape in the form of a 12-D
real valued fingerprint. As a result, the calculation procedure is
rapid and once calculated shape similarities can be evaluated rapidly
via a simple distance calculation. It should also be noted that shape
indices based on the molecular topology have also been described
\cite{Kier:1985aa}, which are very rapid to calculate (somewhat
analogous to the VSA descriptors) but do not directly represent the
molecular shape and instead infer it from topological connectivity.

The last category consists of QM descriptors. These
descriptors can also be applied to problems such as QSAR models
(though of course the time complexity of QM derived descriptors do not
make them amenable for large scale virtual screening
applications). These descriptors can be evaluated at various levels of
theory. For example, partial charges, HOMO and LUMO energies are
commonly evaluated using semi-empirical methods, while Eroglu and
T\"{u}rkmen \cite{Eroglu:2007uq} described DFT-derived descriptors
such as electronegativity, hardness and ionization potential (though
Puzyn et al question the need for such a high level of theory
\cite{Puzyn:2008fk}).

@@

Every descriptor implements an interface allowing them to
return values in a uniform manner, viz., via a \texttt{DescriptorValue}
object (see Figure~\ref{fig:cdkapi}). This class can represents the result
of a descriptor calculation and includes both the descriptor values
(via a \texttt{IDescriptorResult} object) as well as meta-data on the
descriptor itself via a \texttt{DescriptorSpecification} object (see
below for more details). The descriptor classes are designed such that
in the event of a calculation error (invalid, missing or nonsensical
input), null values are returned, rather than throwing an
exception. This is useful during automated descriptor calculation,
where one desires a ``rectangular'' descriptor matrix, from which one
could filter out the undefined descriptors. While the user can pick
and choose specific descriptors to calculate, the API provides a class
(\texttt{DescriptorEngine}) to specifically evaluate all or a subset
of descriptors for a molecule at one go.  Currently the descriptor
implementations are not threadsafe, making them unsuitable for
parallelization. This is a drawback since many of the descriptors can
be slow. For example, the $\chi$ descriptors \cite{Kier:1976aa} are
implemented using SMARTS matching to identify the various paths and
chains described by Kier \textit{et al.}, rather than explicitly walking the
molecular graph.

@@

Given these challenges it is imperative that descriptor tools provide
access to version information. This allows the user to provide an
exact specification of how the descriptor was calculated. However, one
aspect that has not been standardized across different implementations
is the naming scheme. In other words, tools that evaluate the same
descriptor may name them in a similar but not exactly identical
manner. This makes automated comparisons and merging of descriptors
from different sources problematic. To address the CDK has developed
the concept of a ``descriptor specification'' which is associated with
a descriptor value and includes information on the vendor, descriptor
title and a reference to an entry in a descriptor dictionary that
contains more details of the descriptor in RDF. While the extra
descriptor metadata is currently used to implement a simple
classification scheme, it is conceivable that in the future the
specification approach could be adopted by multiple vendors, allowing
for automated reasoning over descriptor implementations.

@@

%% See http://code.google.com/p/rdkit/w/list for RDKit
%% See
%% http://www.ra.cs.uni-tuebingen.de/software/joelib/tutorial/descriptors/fingerprints.html for JOELib
%% http://www.ra.cs.uni-tuebingen.de/software/joelib/tutorial/applications/descriptorcalculation.html for descs
\ctable[caption={A summary of descriptor implementations covering Open
  Source and commercial cases.}, cap={}, label={tab:impl}]
{llllp{0.3\linewidth}p{0.2\linewidth}}
{}
{
\FL
Name & License & Toolkit & Application & Descriptors & Fingerprints
\ML
CDK & LGPL & Yes & Yes &
\raggedright{Constitutional, topological, geometric, electronic and hybrid} &
  \raggedright{MACCS, PubChem, Kier \& Hall Estate keys. Hashed linear
  and circular} \NN
CODESSA & Commercial & No & Yes & \raggedright{Constitutional,
  topological, geometric, electronic, QM, thermodynamic} & \NN
Dragon & Commercial & No & Yes & \raggedright{Extensive list
of all classes of descriptor types and implements all those described
in Ref.~\citen{Todeschini:2002aa}} &  \NN
JChem & Commercial & Yes & Yes & \raggedright{Consititutional,
  geometric, topological, electronic, hybrid} & \raggedright{Hashed,
  circular, ECFP/FCFP} \NN
JOELib & GPL & Yes & Yes & \raggedright{} &
\raggedright{Keys from Xue et al\ \cite{Godden:2000qf}} \NN
Molconn-Z & Commercial & No &  Yes & topological &  \NN
OpenBabel & GPL & Yes & & \raggedright{} & \raggedright{} \NN
PowerMV \cite{Liu:2005ab} & GPL &  & Yes & \raggedright{constitutional, atom pairs,
  Burden numbers, pharmacophoric} & \raggedright{} \NN
RDKit & New BSD & Yes & & \raggedright{Constitutional, topological,
  EState, partial charges and VSA combining EState, partial charges
  and MR} & MACCS, hashed, atom pairs, topological
torsions 
\LL
}

@@

\ctable[caption={UML diagram of the CDK descriptor API in version 1.4.},cap={},label={fig:cdkapi},botcap,figure]
{c}
{}
{
  \includegraphics[width=\linewidth]{iDescriptor.pdf}
}

>>>>

Methods that aim to predict chemical and biological properties
generally require a numerical description of chemical structures. Such
numerical forms range from a set of 3D coordinates which coupled with
appropriate atom types, is sufficient for methods such as quantum
mechanical (QM) approaches and docking to more abstract numerical
descriptions derived from 2D or 3D representations which can be useful
in statistical approaches. It is now possible to evaluate thousands of
numerical descriptors of chemical structure. As will be discussed
later, many of these descriptors are closely related or capture the
same information, allowing one to be substituted for another. The
choice of descriptors is a well-known problem and given a large
collection of them, approaches to identify a suitable subset have been
discussed extensively in the
literature~\cite{Miller:2002aa,Kohavi:1997gf}.

In addition to there being many possible descriptors defined in the
literature, there are also multiple implementations of a give
descriptor. These implementations are available in the form of
libraries (which require one to write a program to use them) or
complete applications (graphical user interface (GUI) or command
line). As a result, not only must one choose one or more descriptors
that are relevant to the problem at hand, but one must be concerned
about how they are calculated and whether such a calculation can be
reproduced across different implementations of those descriptors. It
is easy to understand \emph{why} two implementations of the same
descriptor can lead to different results. The primary reasons being
differences in the chemistry model of the framework or toolkit used to
implement the descriptor. For example, a descriptor that calculates
the number of aromatic atoms may be implemented using two toolkits
with differing aromaticity models, and hence it is possible that the
values generated by the two implementations will differ. Other sources
of differences include parameters that may be involved in the
descriptor calculation and reference data values (such as atomic
radii, electronegativity values) that are employed during descriptor
calculation. While most implementations will employ the same data
sources for standard concepts (e.g., atomic weights), slight
differences in these types of input data can lead to differences in
the final descriptor value~\cite{Guha:2006ac}. As a result, in most
cases, two implementations of a descriptor do not usually give the
exact same value, though they are usually quite similar. Explicitly
explaining the differences may or may not be possible (it is usually
more difficult in the cases of commercial implementations for which
source code is not available)~\cite{Spjuth2010}.

@@

\subsection{Numerical descriptors}
\label{sec:numer-descr}

@@

Another important class of geometrical descriptors are those that
describe molecular shape. This property is a key factor in ligand-receptor interactions and is
an important approach to virtual screening. A variety of descriptors
have been devised to characterize molecular shape. The use of
overlapping atom centered Gaussians to represent molecular shape
\cite{Grant:1995aa} represents a significant computational advantage
over the traditional analytical approach of intersecting hard spheres,
supporting extremely rapid shape comparisons \cite{Grant:1999vn}. An
alternative approach described by Ritchie et al \cite{Ritchie:1999kx}
employed a spherical harmonic based representation of surfaces, that
in contrast to the Gaussian approach allows one to generate
characterizations of the surface and hence the shape at varying
resolutions. Another approach to multi-resolution shape
representations is the use of $\alpha$ shapes
\cite{Edelsbrunner:1994aa}, which represent a parameterization of the
convex hull derived from the molecular coordinates. This approach has
been employed to characterize shape and volume \cite{Wilson:2009ys}
though it is more computationally intensive than the methods mentioned
previously. In addition to functional representations of molecular
shape, a number of simpler formulations have been defined. For
example, the ``shape signature'' approach described by Zauhar et al
\cite{Zauhar:2003fk} is a novel application of a ray tracing method
that generates a description of the molecular shape. Ballester et al
\cite{Ballester:2007aa} described the Ultrafast Shape Recognition
(USR) approach that characterized molecular shape in terms of the
distribution of distances between the atoms and a set of pre-defined
points. This procedure characterizes the shape in the form of a 12-D
real valued fingerprint. As a result, the calculation procedure is
rapid and once calculated shape similarities can be evaluated quickly
via a simple distance calculation. It should also be noted that shape
indices based on the molecular topology have also been described
\cite{Kier:1985aa}, which are very rapid to calculate (somewhat
analogous to the VSA descriptors) but do not directly represent the
molecular shape and instead infer it from topological connectivity.

The last category consists of QM descriptors. These
descriptors can also be applied to problems such as QSAR models
(though of course the time complexity of QM derived descriptors do not
make them amenable for large scale virtual screening
applications). These descriptors can be evaluated at various levels of
theory. For example, partial charges, HOMO and LUMO energies are
commonly evaluated using semi-empirical methods, while Eroglu and
T\"{u}rkmen \cite{Eroglu:2007uq} described DFT-derived descriptors
such as electronegativity, hardness and ionization potential (though
Puzyn et al question the need for such a high level of theory
\cite{Puzyn:2008fk}). Sullivan et al \cite{Sullivan:2000fk} employed
frontier orbital theory and DFT to generate quantum mechanical analogs
of Hammett substituent constants \cite{Hammett:1937aa} and showed that
they agreed well with empirically derived constants.

@@

Every descriptor class implements an interface allowing them to return
values in a uniform manner, viz., via a \texttt{DescriptorValue}
object. This class can represents the result of a descriptor
calculation and includes both the descriptor values (via a
\texttt{IDescriptorResult} object) as well as meta-data on the
descriptor itself via a \texttt{DescriptorSpecification} object (see
below for more details). The descriptor classes are designed such that
in the event of a calculation error (invalid, missing or nonsensical
input), null values are returned, rather than throwing an
exception. This is useful during automated descriptor calculation,
where one desires a ``rectangular'' descriptor matrix, from which one
could filter out the undefined descriptors. While the user can pick
and choose specific descriptors to calculate, the API provides a class
(\texttt{DescriptorEngine}) to specifically evaluate all or a subset
of descriptors for a molecule at one go.  Currently the descriptor
implementations are not threadsafe, making them unsuitable for
parallelization. This is a drawback since many of the descriptors can
be slow. For example, the $\chi$ descriptors \cite{Kier:1976aa} are
implemented using SMARTS matching to identify the various paths and
chains described by Kier \textit{et al.}, rather than explicitly
walking the molecular graph.

@@

Given these challenges it is imperative that descriptor tools provide
access to version information. This allows the user to provide an
exact specification of how the descriptor was calculated. However, one
aspect that has not been standardized across different implementations
is the naming scheme. In other words, tools that evaluate the same
descriptor may name them in a similar but not exactly identical
manner. This makes automated comparisons and merging of descriptors
from different sources problematic. To address this, the CDK has developed
the concept of a ``descriptor specification'' which is associated with
a descriptor value and includes information on the vendor, descriptor
title and a reference to an entry in a descriptor dictionary that
contains more details of the descriptor in RDF. While the extra
descriptor metadata is currently used to implement a simple
classification scheme, it is conceivable that in the future the
specification approach could be adopted by multiple vendors, allowing
for automated reasoning over descriptor implementations.

@@

%% See http://code.google.com/p/rdkit/w/list for RDKit
%% See
%% http://www.ra.cs.uni-tuebingen.de/software/joelib/tutorial/descriptors/fingerprints.html for JOELib
%% http://www.ra.cs.uni-tuebingen.de/software/joelib/tutorial/applications/descriptorcalculation.html for descs
\ctable[caption={A summary of descriptor implementations covering Open
  Source and commercial cases.}, cap={}, label={tab:impl}]
{llllp{0.3\linewidth}p{0.2\linewidth}}
{}
{
\FL
Name & License & Toolkit & Application & Descriptors & Fingerprints
\ML
CDK & LGPL & Yes & Yes &
\raggedright{Constitutional, topological, geometric, electronic and hybrid} &
  \raggedright{MACCS, PubChem, Kier \& Hall Estate keys. Hashed linear
  and circular} \NN
CODESSA & Commercial & No & Yes & \raggedright{Constitutional,
  topological, geometric, electronic, QM, thermodynamic} & \NN
Dragon & Commercial & No & Yes & \raggedright{Extensive list
of all classes of descriptor types and implements all those described
in Ref.~\citen{Todeschini:2002aa}} &  \NN
JChem & Commercial & Yes & Yes & \raggedright{Consititutional,
  geometric, topological, electronic, hybrid} & \raggedright{Hashed,
  circular, ECFP/FCFP} \NN
JOELib & GPL & Yes & Yes & \raggedright{} &
\raggedright{Keys from Xue et al\ \cite{Godden:2000qf}} \NN
Molconn-Z & Commercial & No &  Yes & topological &  \NN
OpenBabel & GPL & Yes & & \raggedright{constitutional} &
\raggedright{MACCS, hashed} \NN
PowerMV \cite{Liu:2005ab} & GPL &  & Yes & \raggedright{constitutional, atom pairs,
  Burden numbers, pharmacophoric} & \raggedright{} \NN
RDKit & New BSD & Yes & & \raggedright{Constitutional, topological,
  EState, partial charges and VSA combining EState, partial charges
  and MR} & MACCS, hashed, atom pairs, topological
torsions 
\LL
}

@@

% \ctable[caption={UML diagram of the CDK descriptor API in version 1.4.},cap={},label={fig:cdkapi},botcap,figure]
% {c}
% {}
% {
%   \includegraphics[width=\linewidth]{iDescriptor.pdf}
% }
